Automated discovery \\ of interpretable gravitational-wave population models

### 2022-07-17 src/tex/ms.tex

\begin{abstract}
We present an automatic approach to discover analytic population models for gravitational-wave (GW) events from data.
As more gravitational-wave (GW) events are detected, flexible models such as Gaussian Mixture Models have become more important in fitting the distribution of GW properties due to their expressivity.
However, flexible models come with many parameters that lack physical motivation, making interpreting the implication of these models challenging.
In this work, we demonstrate symbolic regression can complement flexible models by distilling the posterior predictive distribution of such flexible models into interpretable analytic expressions.
We recover common GW population models such as a power-law-plus-Gaussian, and find a new empirical population model which combines accuracy and simplicity.
This demonstrates a promising pathway to automatically discover interpretable population models in the ever-growing GW catalog, which can potentially be applied to other astrophysical phenomena.
\end{abstract}

@@

Alternatively, there have been studies using much more flexible models such as a Gaussian Mixture Model (GMM) to analyze the catalog \cite{Tiwari:2020vym}.
While such a flexible model is powerful in fitting complex datasets,
it is difficult to interpret the physical meaning of the fitted parameters.
There are two reasons why such flexible models are hard to interpret:
first, these models often have many parameters;
in GWTC3, the GMM has over 100 parameters.
With so many degrees of freedom, it is non-trivial to analyze contributions from each mode and attempt to disentangle meaningful patterns.
Second, flexible models often employ bases that are not physically motivated.
For example, the GMM models has 11 independent Gaussian components in the mass-spin space.
While one or two Gaussians in mass or spin space alone can be physically motivated,
the rest of the components are included in the model jsut to fit for unexpected features.
Therefore, interpreting the importance of each component could be complicated by the components that are not physically motiviated.

@@

While most ML algorithms optimize a fixed set number of real parameters,
Symbolic regression is a machine learning (ML) algorithm which searches for analytic expressions which optimize some loss function.
The search typically employs a ``genetic algorithm'', which arranges symbols in a tree, evaluates the resultant expression, and repeats this process until a particular goal is met.
Since models for physical phenomena are often specified in a language of concise mathematical expressions, symbolic regression can produce interpretable models compared to traditional ML, as the ``trained model'' generated by symbolic regression is in the same \textit{language} as existing physical models (i.e., analytic expressions).
Therefore, symbolic regression can produce insights for science in a way that traditional ML cannot.

@@

Internal to \pysr, a multi-objective optimization problem is solved: the evolution seeks to find expressions which are as accurate as possible, while being as simple as possible.
Formally, for our setting, \pysr solves the following optimization problem:
\begin{align}
    s_c &= 
        \mathop{\arg\min}_{
            \tiny
            \begin{array}{c}s\in\mathcal{S}\\ \abs{s} = c \end{array}
        }
        \mathcal{L}(s),
    \quad\text{for each } c\in \mathcal{C},
\end{align}
for the loss function $\mathcal{L}(s) = \sum_i \ell(s(\mathbf{x}_i), y_i),$
where $s$ is an expression, 
$\mathcal{S}$ is a space of possible expressions considered,
$\abs{s}$ is the complexity of an expression $s$,
$\ell$ is the loss function, and $\{(x_i, y_i)\}$ is the dataset
for $x_i$ each input and $y_i$ the label.
The most accurate expression at each $c$, denoted as $s_c$, is computed for a given set range of $c$, which we choose as $c\in\mathcal{C}\equiv \{1, \ldots, 50\}$.
Complexity is defined here
as the number of symbols in the expression (including repeats),
and $\ell$ is defined as $(\hat{y}_i, y_i) = \qty(\hat{y}_i - y_i)^2$.

Next, given the set of the best expressions at each complexity $\mathcal{S}^\ast = \{s_c\}_{c\in\{1,\ldots,80\}}$,
Next, we can construct the \textit{Pareto front}, defined as:
\begin{equation}
    \hat{\mathcal{S}}^\ast = \{s_c \ |\ \mathcal{L}(s_c) < \mathcal{L}(s_q) \forall q < c \},
\end{equation}
such that each remaining expression is more accurate than all simpler expressions.
Using this, we can define the \textit{score} of an expression as the slope of the Pareto front, equal to: $\text{score} = -\Delta(\log \mathcal{L}(s_c))/\Delta c,$
which is a heuristic for the quality of a particular equation, used in the symbolic regression literature \citep[see, e.g.,][]{schmidtDistillingFreeFormNatural2009,cranmerDiscoveringSymbolicModels2020}.

@@

We show the workflow of this work in figure \ref{fig:Gaussian_m1_fit}.
On the leftmost panel, we show the posterior density in primary mass for every event in GWTC3.
In the left-middle panel, we show a GMM fitted to the data given by \cite{Tiwari:2020vym}.
Given the \ppd shown in the middle panel, we use symbolic regression to extract effective descriptions of the data. 
As we increase the complexity of the equation, it is natural to expect that the equation should fit the data better than a lower complexity equation since it has additional free parameters.
At the same time, an equation with higher complexity could either be overfitting or difficult to interpret.
\pysr uses a score (see \citealt{cranmerDiscoveringSymbolicModels2020} for details) as a metric normalized to complexity to show how well an equation is fitting the data.
At each complexity, \pysr outputs the equation with the lowest loss.
To select the equations shown in figure \ref{fig:Gaussian_m1_fit}, we first make an accuracy cut based on the loss of the function such that all expressions have a loss lower than $0.3$, so the selected equation would still be a good fit to the data.
We then consider the expressions with the highest score among three ranges: $c\in [1, 20]$,
$c\in[20, 35]$, and $c\in[35, 50]$.
These ranges are set to simply give us a variety of complexity expressions to study.
The equations selected under these criteria are tabulated in table \ref{tab:equations}.

@@

Compared to the lowest complexity equation, the middle complexity (25) equation instead prefers a lower normalization of the exponential function,
but account for the bump around $30\ M_{\odot}$ with a Gaussian component,
which is proposed to be driven by a pile-up from pulsational pair-instability supernovae \cite{Talbot:2018cva}.
This equation is similar to one of the simplest phenomenological models fitted to GWTC3, which is a power law mixed with a Gaussian bump.
The difference between this equation and the one presented in GWTC3 is this equation has an additional bump at $m_{1}\sim 10\ M_{\odot}$,
which is also present in the lower complexity equation.
This suggests the first bump at lower mass has more statistical significance than the second bump.
Although the middle complexity equation is more complex, the loss is about one order of magnitude lower than the next best one with lower complexity (16).
As a result, the score of this equation is higher than the lower complexity equation, making it a favorable equation. This means the second Gaussian component is needed in order to describe the distribution of GWs.

At the highest complexity (48), one more Gaussian component is added to the equation in order to account for the tapering at the higher mass end. 
Interestingly, the new component added in this equation does not try to fit the wiggles in between the first bump and the second bump.
The significance of any features other than the two bumps and the continuum is still an open question.
Our result hints the wiggles between the bumps seem to be less significant than the extra tapering of the continuum.
In addition, none of the equations prefers to fit the extra hump before the $10\ M_{\odot}$ peak.
This could mean the little hump could be an artifact in the Gaussian mixture model instead of a truly statistically significant feature.
However, the loss of this equation is similar to the middle complexity (25) equation, at the same time being almost twice as complex. As a result, the score is much lower, which means the tapering is not necessarily needed.

@@

The main message of this work is that by combining flexible models with symbolic regression, we can discover interpretable equations that can describe the increasingly complex GW catalog.
Compared to traditional methods such as phenomenological modeling, which relies on an individual's insight,
our pipeline requires minimal human intervene.
Our method is not a model selection method by itself, but a pipeline to discover models for downstream tasks such as model selection and parameter estimation.
Once we obtain a set of equations, we can plug the expressions in a standard HBA pipeline to estimate the parameters of the model and select between models.
This allows the community to discover interpretable patterns in the GW catalog in a data-driven instead of model-driven manner.
In this sense, our method is more scalable to more complex and higher-dimensional datasets.

In this work, we focus on analyzing only the primary mass distribution.
Our method is trivially generalizable to higher-dimensional data, such as the joint space of both the primary mass and the secondary mass.
So far there are relatively few phenomenological models which can capture correlations between observables,
mainly due to the difficulty of manually constructing reasonable models to account for these correlations.
At the same time, the most advanced flexible model in the literature is an n-dimensional histogram, which is hard to interpret.
Our method can be a natural way to explore and to explain these correlations.

Although we focus on GW in this work, the method can be easily extended to other population analysis.
For example, given the recently release Gaia DR3 data \cite{Collaboration2016TheGM}, we can apply our method to understanding the population of stars without much augmentation to the pipeline.
As size of data sets grows larger in astronomy, this work provides a simple and scalable pathway to uncover trends and relations in astrophysical catalogs.

@@

\clearpage
\appendix
\section{Operator constraints}
\label{sec:constraints}

>>>>

\begin{abstract}
We present an automatic approach to discover analytic population models for gravitational-wave (GW) events from data.
As more gravitational-wave (GW) events are detected, flexible models such as Gaussian Mixture Models have become more important in fitting the distribution of GW properties due to their expressivity.
However, flexible models come with many parameters that lack physical motivation, making interpreting the implication of these models challenging.
In this work, we demonstrate symbolic regression can complement flexible models by distilling the posterior predictive distribution of such flexible models into interpretable analytic expressions.
We recover common GW population models such as a power-law-plus-Gaussian, and find a new empirical population model which combines accuracy and simplicity.
This demonstrates a strategy to automatically discover interpretable population models in the ever-growing GW catalog, which can potentially be applied to other astrophysical phenomena.
\end{abstract}

@@

Alternatively, there exist studies using much more flexible models such as  Gaussian Mixture Models (GMMs) to analyze the catalog \cite{Tiwari:2020vym}.
While such a flexible model is powerful in fitting complex datasets,
it is difficult to interpret the physical meaning of the fitted parameters.
There stand two core reasons why such flexible models are hard to interpret:
first, these models often have many parameters;
in GWTC3, the GMM has over 100 parameters.
With so many degrees of freedom, it is non-trivial to analyze contributions from each mode and attempt to disentangle meaningful patterns.
Second, flexible models often employ bases that are not physically motivated.
For example, the GMM models has 11 independent Gaussian components in the mass-spin space.
While one or two Gaussians in mass or spin space alone can be physically motivated,
the additional components are included in the model fit for unexpected features.
Therefore, interpreting the importance of each component could be complicated by the components that are not physically motivated.

@@

While most ML algorithms optimize a fixed set number of real parameters,
symbolic regression is a machine learning (ML) algorithm which searches for analytic expressions which optimize some loss function.
The search typically employs a ``genetic algorithm,’’ which arranges symbols in a tree, evaluates the resultant expression, and repeats this process until a particular target is met.
Since models for physical phenomena are often specified in a language of concise mathematical expressions, symbolic regression can produce interpretable models compared to traditional ML, as the ``trained model'' generated by symbolic regression is in the same \textit{language} as existing physical models (i.e., analytic expressions).
Therefore, symbolic regression can produce insights for science in a way that traditional ML cannot.

@@

Internal to \pysr, a multi-objective optimization problem is solved: the evolution seeks to find expressions which are as accurate as possible, while being as simple as possible.
Formally, for our setting, \pysr solves the following optimization problem:
\begin{align}
    s_c &= 
        \mathop{\arg\min}_{
            \tiny
            \begin{array}{c}s\in\mathcal{S}\\ \abs{s} = c \end{array}
        }
        \mathcal{L}(s),
    \quad\text{for each } c\in \mathcal{C},
\end{align}
for the loss function $\mathcal{L}(s) = \sum_i \ell(s(\mathbf{x}_i), y_i),$
where $s$ is an expression, 
$\mathcal{S}$ is a space of possible expressions considered,
$\abs{s}$ is the complexity of an expression $s$,
$\ell$ is the loss function, and $\{(x_i, y_i)\}$ is the dataset
for $x_i$ each input and $y_i$ the label.
The most accurate expression at each $c$, denoted as $s_c$, is computed for a given set range of $c$, which we choose as $c\in\mathcal{C}\equiv \{1, \ldots, 50\}$.
Complexity is defined here
as the number of symbols in the expression (including repeats),
and $\ell$ as $\ell(\hat{y}_i, y_i) = \qty(\hat{y}_i - y_i)^2$.

Next, given the set of the best expressions at each complexity $\mathcal{S}^\ast = \{s_c\}_{c\in\{1,\ldots,80\}}$,
Next, we can construct the \textit{Pareto front}:
\begin{equation}
    \hat{\mathcal{S}}^\ast = \{s_c \ |\ \mathcal{L}(s_c) < \mathcal{L}(s_q) \forall q < c \},
\end{equation}
such that each remaining expression is more accurate than all simpler expressions.
Using this, we can define the \textit{score} of an expression as the slope of the Pareto front, equal to: $\text{score} = -\Delta(\log \mathcal{L}(s_c))/\Delta c,$
which is a heuristic for the quality of a particular equation, used in the symbolic regression literature \citep[see, e.g.,][]{schmidtDistillingFreeFormNatural2009,cranmerDiscoveringSymbolicModels2020}.

@@

We show the workflow of this work in figure \ref{fig:Gaussian_m1_fit}.
On the leftmost panel, we show the posterior density in primary mass for every event in GWTC3.
In the left-middle panel, we show a GMM fitted to the data given by \cite{Tiwari:2020vym}.
Given the \ppd shown in the middle panel, we use symbolic regression to extract effective descriptions of the data. 
As we increase the complexity of the equation, it is natural to expect that the equation should fit the data better than a lower complexity equation since it has additional free parameters.
At the same time, an equation with higher complexity could either be overfitting or difficult to interpret.
\pysr uses a score (see \citealt{cranmerDiscoveringSymbolicModels2020} for details) as a metric normalized to complexity to show how well an equation is fitting the data.
At each complexity, \pysr outputs the equation with the lowest loss.
To select the equations shown in figure \ref{fig:Gaussian_m1_fit}, we first make an accuracy cut based on the loss of the function such that all expressions have a loss lower than $0.3$, so the selected equation would still be a good fit to the data.
We then consider the expressions with the highest score among three ranges: $c\in [1, 20]$,
$c\in[20, 35]$, and $c\in[35, 50]$.
These ranges are set to give us a variety of complexity expressions to study.
The equations selected under these criteria are tabulated in table \ref{tab:equations}.

@@

Compared to the lowest complexity equation, the middle complexity (25) equation instead prefers a lower normalization of the exponential function,
but account for the bump around $30\ M_{\odot}$ with a Gaussian component,
which is proposed to be driven by a pile-up from pulsational pair-instability supernovae \cite{Talbot:2018cva}.
This equation is similar to one of the simplest phenomenological models fitted to GWTC3, which is a power law mixed with a Gaussian bump.
The difference between this equation and the one presented in GWTC3 is this equation has an additional bump at $m_{1}\sim 10\ M_{\odot}$,
which is also present in the lower complexity equation.
This suggests the first bump at lower mass has more statistical significance than the second bump.
Although the middle complexity equation is more complex, the loss is about one order of magnitude lower than the next best one with lower complexity (16).
As a result, the score of this equation is higher than the lower complexity equation, making it a favorable equation.

At the highest complexity (48), one more Gaussian component is added to the equation account for the tapering at the higher mass end. 
Interestingly, the new component added in this equation does not try to fit the wiggles in between the first bump and the second bump.
The significance of any features other than the two bumps and the continuum is still an open question.
Our result hints the wiggles between the bumps seem to be less significant than the extra tapering of the continuum.
In addition, none of the equations prefers to fit the extra hump before the $10\ M_{\odot}$ peak.
This could mean the little hump could be an artifact in the Gaussian mixture model instead of a truly statistically significant feature.
However, the loss of this equation is similar to the middle complexity (25) equation, at the same time being almost twice as complex. As a result, the score is much lower, which means the tapering is not necessarily needed.

@@

The main message of this work is that by combining flexible models with symbolic regression, we can discover interpretable equations that can describe the increasingly complex GW catalog.
Compared to traditional methods such as phenomenological modeling, which relies on an individual scientist's insight,
our pipeline requires minimal human intervention.
 This technique is not a model selection method by itself, but a pipeline to discover models for downstream tasks such as model selection and parameter estimation.
Once we obtain a set of equations, we can insert the expressions to a standard HBA pipeline to estimate the parameters of the model and select between models.
This allows the community to discover interpretable patterns in the GW catalog in a data-driven instead of hand-crafted model-driven manner.
In this sense, our method is more scalable to more complex and higher-dimensional datasets.

In this work, we focus on analyzing only the primary mass distribution.
Our method is trivially generalizable to higher-dimensional data, such as the joint space of both the primary mass and the secondary mass.
So far there are relatively few phenomenological models which can capture correlations between observables,
mainly due to the difficulty of manually constructing reasonable models to account for these correlations.
At the same time, the most advanced flexible model in the literature is an n-dimensional histogram, which is hard to interpret.
Our method provides a natural way to explore and to explain these correlations.

Although we focus on GW in this work, this method can be easily extended to other population analysis problems.
For example, given the recently released Gaia DR3 dataset \cite{Collaboration2016TheGM}, it might be possible to apply this method to understanding various subpopulations of  stars in the Milky Way.
As size of datasets grows larger in astronomy, this work provides a simple and scalable pathway to uncover trends and relations in astrophysical catalogs.

@@

\clearpage
\appendix
\section{Operator Constraints}
\label{sec:constraints}
